Modal Abstractions for Virtualizing Memory Addresses

### 2023-04-13 programlogic.tex
compiles again
starting to clean up section 4

\newcommand{\sumapaces}[2]{
  \ownGhost\gammaPreds{\authfrag{\singletonMap{#1}{#2}}}
\newcommand{\ptableabswalk}[1]{\mathcal{A}\textsf{bsPTableWalk}(#1)}
\newcommand{\ptablestore}{\theta}
Our reasoning principles derive from the following perspectives on the virtual-memory abstraction:
\begin{enumerate}
\item \textit{logical representation of addressing}: we present physical memory addressing as separation-logic points-to relations.
\item \textit{sharing}: the physical location backing a virtual address is located (during a page table walk) through a 
      set of physical page-table (L4-L1 page-tables) acceses that are shared amongst different virtual addresses (specifically,
      those on the same page of memory). This sharing imposes constraints on giving a definition to the virtual-address 
      in terms of physical (L4-L1) page-table memory accesses
\item \textit{context-agnostic-resources}: each virtual address is valid under a certain address-space, but it does not represent this \textit{knowledge} of its address-space. 
\item \textit{address-spaces as modal contexts}: we call these context-dependent\todo{Colin switched away from `agnostic', which is the opposite!} resources (virtual-address pointstos) as \textit{context-resource}, and the context identfied by a unique \textit{namespace}, and  defining the truth for these context resources as \text{modal-context}
\item \textit{updating inter/intra address-spaces}: we present logical abstractions to enable allocating/updating pages in the existence of \textit{sharing}  
\item \textit{address-space switch as changing the "World" of truth}: switching from one address-space to another logically becomes introduction of \textit{context-resources} to the \textit{modal-context} under which they are valid, and loading \textit{context-resources} that are valid under the \textit{currently} loaded address-space
\end{enumerate}
Based on these perspectives, we coneptualize truth in an address-space through the contingency it exhibits: 
\textit{assertion P holds only relative to a certain address space}. 
Therefore, as a first step, we work in a pointwise lifting of Iris's standard BI proposition,
essentially working with separation logic assertions indexed by a page table root as a $\mathcal{W}_{64}$
(Figure \ref{fig:vprop}), which we call $\textsf{vProp }\Sigma$; this is the (\textsc{Coq}) type of assertions in our logic.
\begin{figure}[t]
\begin{lstlisting}[language=Coq]
  Definition vProp  $\Sigma$ : bi := word 64 -b> iPropI  $\Sigma$.
  \end{lstlisting}
\caption{$\textsf{vProp }\Sigma$: Root-Address Indexed Address-Space Proposition}
  \label{fig:vprop}
%\subsection{Machine State under Address Translation}
%\label{sec:selectedinstrsemantics}
%Althought we give the complete set of operational semantics rule in \sref{appendix:movops}, It is worth building the intuition on the the way some of these rules bahave in the context of address translation. In fact $\readlval\maddr{\storememstar\crval}\locsf$ is unfoled into multiple physical memory lookups for the final page address retrieval -- i.e. address translation traversal as shown in Figure \ref{fig:pagetables}:
%\begin{itemize}
%\item top-level-address translation: with the given root address (64-bit $\crval$) of the address space, performs the address translation, handling the first level (to get the PML4 entry) itself. The next level table address is computed with the fetched PML4 offset value which exhibits itself as 9-bit offset in $\kw{maddr}$. \mytodo{iso: put move}
%\item translating from PML4 entry: performs the second level of address translation, to retrieve starting at the PML4 table entry, and interprets the PML4 entry that references a Page-Directory-Pointer Table (PDPT). We obtain the PDPT offset which exhibits itself as 9 bit offset in $\kw{maddr}$ to obtain the address of the next level page directory table (PDT) \mytodo{iso: put move}
%\item translating from PD entry: performs the third level of address translation, to retrieve starting at the PDP table entry, and interprets a PDPTE that references a PD table. Likewise, we obtain the PD offset which exhibits itself as 9 bit offset in $\kw{maddr}$ to obtain the address of the next level page directory table (PT) \mytodo{iso:put move }
%\item translate from PT entry: performs final level of address translation, starting from the PT entry, with a given 12 bit page offset, we can compute the physical address referencing $\locsf$ \mytodo{ismo: put mov}
%\end{itemize}
\subsection{Points-To Assertions}
\label{sec:pointsto}
\todo[inline]{Colin says: Current limit of \emph{thorough} revisions}
\todo[inline,color=red]{Colin says: This section at a glance looks on the right track, but none of the assertions are actually
written as vProps, they're all still iProps. I can fix that as I go}
\todo[inline]{Colin says: There's really no need to discuss the 52-12 split in the physical pointsto definition.
That level of detail of the operational semantics \emph{is} too detailed for the paper (but should be mentioned in the
artifact docs). I also don't know what the `ent' and `rtv' and similar are in those split physical pointstos}
We abstract physical memory addressing and registers naming with well-known separation logic assertion \textit{points-to} assertions.
\begin{enumerate}
\item Physical address points-to, $\pfpointsto\locsf\vpts\qfrac\ppts$
\todo[inline,color=red]{Colin says: This syntax is what most readers would expect for physical poinst-tos, but it's
inconsistent with the split 52+12  stuff later... Whatever we do, we need to be consistent}
\item Register points-to, $\pfpointsto\rg\rv\qfrac\rpts$
\end{enumerate}

@@

@@

\todo[inline,color=red]{Colin says: Ismail, I think you accidentally deleted discussion of physical points-tos here?}

\todo[inline,color=red]{Colin says: Why are we talking about assertions for address translations here? It seems like we should
\todo[inelin]{Colin says: There's no explanation of the split 52+12 \emph{notation} here. If we added such an explanation
(maybe to where physical points-tos are introduced?) and moved this discussion to \ref{sec:overly-restrictive}, this is otherwise
a nice explanation}
discuss physical memory pointsto assertions in general first, then talk about this in \ref{sec:overly-restrictive}}
Given the definition of physical page-pointsto assertion and the root address of virtual-address space as shown in Figure \ref{fig:pagetable}, one can build the physical address-translation for a virtual address (e.g. \textsf{va}) via abstracting the L4-L1 table traversal as the following:
\begin{itemize}
  \item Level-4 Translation (L4): Performs the first level address translation to get the L4 entry (L4E) in Figure \ref{fig:pagetables} by using the masks l4M52 and l4off with \textsf{rtv} virtual base address to get the starting address of the L4 table (L4T), and locate the entry amongst 512 (q1) ones respectively.
    \begin{lstlisting}[language=Coq]
      $\hbox{(\TirNameStyle{{L4translate}})} \quad$ (l4M52 maddr rtv) \$\sim$\ (l4off maddr rtv) $\mapsto_{a}$ {q1} l4e 
    \end{lstlisting}
 \item Level-3 Translation (L3): Performs the second level address translation to get the L3 entry (L3E) in Figure \ref{fig:pagetables} by using the masks l3M52 and l3off with L4 entry (l4e) to get the starting address of the L3 table (L3T), and locate the entry amongst 512 ones (q2) respectively.
    \begin{lstlisting}[language=Coq]
    $\hbox{(\TirNameStyle{{L3translate}})}$ (l3M52 maddr l4e) \$\sim$\ (l3off maddr l4e)  $\mapsto_{a}$ {q2} l3e
    \end{lstlisting}
  \item Level-2 Translation (L2): Performs the second level address translation to get the L2 entry (L2E) in Figure \ref{fig:pagetables} by using the masks l2M52 and l2off with L3 entry (l3e) to get the starting address of the L2 table (L2T), and locate the entry amongst 512 ones (q3) respectively.
\begin{lstlisting}[language=Coq]
    $\hbox{(\TirNameStyle{{L2translate}})}$ (l2M52 maddr l3e) \$\sim$\ (l2off maddr l3e)  $\mapsto_{a}$ {q3} l2e
    \end{lstlisting}
  \item Level-1 Translation (L1): Performs the second level address translation to get the L1 entry (L1E) in Figure \ref{fig:pagetables} by using the masks l1M52 and l1off with L2 entry (l2e) to get the starting address of the L1 table (L1T), and locate the entry amongst ones 512 (q4) respectively.
   \begin{lstlisting}[language=Coq]
    $\hbox{(\TirNameStyle{{L1translate}})}$ (l1M52 maddr l2e) \$\sim$\ (l1off maddr l2e)  $\mapsto_{a}$ {q4} l1e
    \end{lstlisting}
  \item Page Address Level Translation: Final computed physical page-address ($\textsf{addr\_L1}(\vaddr$,$\textsf{l1e}$)) points-to the value stored in the address ($\vpage$).
   \begin{lstlisting}[language=Coq]
    $\hbox{(\TirNameStyle{{PageLevelAccess}})} \qquad$ addr_L1($\vaddr,\textsf{l1e}$) $\mapsto_{p} \vpage$.
    \end{lstlisting} 
\end{itemize}

@@

>>>>

\newcommand{\sumapaces}[2]{
  \ownGhost\gammaPreds{\authfrag{\singletonMap{#1}{#2}}}
}
\newcommand{\ptableabswalk}[1]{\mathcal{A}\textsf{bsPTableWalk}(#1)}
\newcommand{\ptablestore}{\theta}

We derive a program logic (a separation logic) supporting the following stances and constraints:
\begin{enumerate}
\item \textit{address-spaces as modal contexts}: Assertions in our logic are context-dependent,
  in the sense that their truth depends on which address space they are used in, due to the need to support virtual points-to assertions.
\item \textit{sharing}: The physical location backing a virtual address's storage is located (during a page table walk) through a 
      set of physical page-table (L4-L1 page-tables) acceses that are shared amongst different virtual addresses (specifically,
      those on the same page of memory\footnote{of in the case of L2 or higher levels of tables, within a given broader region.}).
      This sharing imposes constraints on definiting points-to assertions
      in terms of physical (L4-L1) page-table memory accesses
\item \textit{context-agnostic-resources}: each virtual address is valid under a certain address-space, 
      but it does not represent this \textit{knowledge} of its address-space. 
      That is, assertions are not explicitly tagged with their address space validity
\item \textit{updating address-space mappings}: We present logical abstractions to enable 
      updating not only pages of typical data in memory, but also page tables themselves.\footnote{Prior work relied on unfolding operational semantics
      to verify page table updates.}
\item \textit{explicitly-modal assertions}: Our logic includes a means to talk about facts being true
      in another address space
\item \textit{address-space switch as changing the "World" of truth}: Switching from one address-space to another logically
      becomes a simultaneous introduction-and-elimination of a pair of modal assertions (for different address spaces)
\end{enumerate}

The first viewpoint, that the truth of an assertion is relative to an address space, has far-reaching consequences.
Therefore, as a first step, we work entirely in a pointwise lifting of Iris's base BI logic,
essentially working with separation logic assertions indexed by a page table root as a $\mathcal{W}_{64}$
(Figure \ref{fig:vprop}), which we call $\textsf{vProp }\Sigma$; this is the (\textsc{Coq}) type of assertions in our logic.
Most constructs in Iris's base logic are defined with respect to any BI-algebra (of \textsc{Coq} type \lstinline|bi|), so automatically
carry over to our derived logic.
However, be must still build up from existing Iris primitives to provide new primitives that depend on the address space --- primarily the notion
of virtual points-to.

\begin{figure}[t]
\begin{lstlisting}[language=Coq]
  Definition vProp  $\Sigma$ : bi := word 64 -b> iPropI  $\Sigma$.
  \end{lstlisting}
\caption{$\textsf{vProp }\Sigma$: Root-Address Indexed Address-Space Proposition}
  \label{fig:vprop}
\end{figure}
%\subsection{Machine State under Address Translation}
%\label{sec:selectedinstrsemantics}
%Althought we give the complete set of operational semantics rule in \sref{appendix:movops}, It is worth building the intuition on the the way some of these rules bahave in the context of address translation. In fact $\readlval\maddr{\storememstar\crval}\locsf$ is unfoled into multiple physical memory lookups for the final page address retrieval -- i.e. address translation traversal as shown in Figure \ref{fig:pagetables}:
%\begin{itemize}
%\item top-level-address translation: with the given root address (64-bit $\crval$) of the address space, performs the address translation, handling the first level (to get the PML4 entry) itself. The next level table address is computed with the fetched PML4 offset value which exhibits itself as 9-bit offset in $\kw{maddr}$. \mytodo{iso: put move}
%\item translating from PML4 entry: performs the second level of address translation, to retrieve starting at the PML4 table entry, and interprets the PML4 entry that references a Page-Directory-Pointer Table (PDPT). We obtain the PDPT offset which exhibits itself as 9 bit offset in $\kw{maddr}$ to obtain the address of the next level page directory table (PDT) \mytodo{iso: put move}
%\item translating from PD entry: performs the third level of address translation, to retrieve starting at the PDP table entry, and interprets a PDPTE that references a PD table. Likewise, we obtain the PD offset which exhibits itself as 9 bit offset in $\kw{maddr}$ to obtain the address of the next level page directory table (PT) \mytodo{iso:put move }
%\item translate from PT entry: performs final level of address translation, starting from the PT entry, with a given 12 bit page offset, we can compute the physical address referencing $\locsf$ \mytodo{ismo: put mov}
%\end{itemize}

\subsection{Base Points-To Assertions}
\label{sec:pointsto}
We build up our context-dependent assertions from two basic types of assertions that do not care about address spaces.
% We abstract physical memory addressing and registers naming with well-known separation logic assertion \textit{points-to} assertions.
% \begin{enumerate}
% \item Physical address points-to, $\pfpointsto\locsf\vpts\qfrac\ppts$
% \todo[inline,color=red]{Colin says: This syntax is what most readers would expect for physical poinst-tos, but it's
% inconsistent with the split 52+12  stuff later... Whatever we do, we need to be consistent}
% \item Register points-to, $\pfpointsto\rg\rv\qfrac\rpts$
% \end{enumerate}

@@

\paragraph{Physical memory  points-to} The soundness proofs for our logic's rules largely center around page-table-walk accesses as in Figure \ref{fig:coq_addr_translation},
 where (inside the helper functions) we need to split a 64-bit word representing entries in these tables into 52-bit segments
which specify the physical page number of another page in the walk, and 12-bit segments containing various control bits.
For this reason we have two notions of physical points-to facts. The primitive notion closest to our machine model is:
\[ \textsf{w1} \ \sim \ \textsf{w2} \mapsto_{\textsf{a}} \; \{\textsf{q}\} \; \textsf{v} \]
In this form, we see that \textsf{w1} (a $\mathcal{W}_{52}$) essentially selects a 4KB page of physical memory,
and \textsf{w2} (a $\mathcal{W}_{12}$) is an offset within that page.
% could be an 52-bits masked address to level 4 table 
% ($\textsf{w1 } =( \textsf{ l4M52 maddr cr3val}) $),
%and, expectedly \textsf{w2} is an address computed by page-offset computation (e.g. $\textsf{l4off maddr cr3val}$). 
From this we can derive a more concise physical points-to when the split is unimportant:
% Giving a raw 64-bits memory pointsto assertion becomes
\[\textsf{w} \mapsto_{\textsf{p}} \{q\} \textsf{ v} \stackrel{\triangle}{=} (\textsf{drop 12}~w) \ \sim \ (\textsf{bottom 12}~w)\mapsto_{\textsf{a}} \textsf{ v} \]

@@

@@

The definition in Figure \ref{fig:strongvirtualpointsto} defines a virtual points-to assertion as
existentially quantifying the physical memory addresses of the individual entries in each page table,
asserting that a corresponding page table walk exists (\lstinline|L4_L1_PointsTo|), and then 
\todo[inline]{snapshot... the last bit of that assertion is weird, but I'll fix. I'll also just remove the fraction from that def since it gets
ignored anyways and we provide a better version shortly}
Given the definition of physical page-pointsto assertion and the root address of virtual-address space as shown in Figure \ref{fig:pagetable}, one can build the physical address-translation for a virtual address (e.g. \textsf{va}) via abstracting the L4-L1 table traversal as the following:
\begin{itemize}
  \item Level-4 Translation (L4): Performs the first level address translation to get the L4 entry (L4E) in Figure \ref{fig:pagetables} by using the masks l4M52 and l4off with \textsf{rtv} virtual base address to get the starting address of the L4 table (L4T), and locate the entry amongst 512 (q1) ones respectively.
    \begin{lstlisting}[language=Coq]
      $\hbox{(\TirNameStyle{{L4translate}})} \quad$ (l4M52 maddr rtv) \$\sim$\ (l4off maddr rtv) $\mapsto_{a}$ {q1} l4e 
    \end{lstlisting}
 \item Level-3 Translation (L3): Performs the second level address translation to get the L3 entry (L3E) in Figure \ref{fig:pagetables} by using the masks l3M52 and l3off with L4 entry (l4e) to get the starting address of the L3 table (L3T), and locate the entry amongst 512 ones (q2) respectively.
    \begin{lstlisting}[language=Coq]
    $\hbox{(\TirNameStyle{{L3translate}})}$ (l3M52 maddr l4e) \$\sim$\ (l3off maddr l4e)  $\mapsto_{a}$ {q2} l3e
    \end{lstlisting}
  \item Level-2 Translation (L2): Performs the second level address translation to get the L2 entry (L2E) in Figure \ref{fig:pagetables} by using the masks l2M52 and l2off with L3 entry (l3e) to get the starting address of the L2 table (L2T), and locate the entry amongst 512 ones (q3) respectively.
\begin{lstlisting}[language=Coq]
    $\hbox{(\TirNameStyle{{L2translate}})}$ (l2M52 maddr l3e) \$\sim$\ (l2off maddr l3e)  $\mapsto_{a}$ {q3} l2e
    \end{lstlisting}
  \item Level-1 Translation (L1): Performs the second level address translation to get the L1 entry (L1E) in Figure \ref{fig:pagetables} by using the masks l1M52 and l1off with L2 entry (l2e) to get the starting address of the L1 table (L1T), and locate the entry amongst ones 512 (q4) respectively.
   \begin{lstlisting}[language=Coq]
    $\hbox{(\TirNameStyle{{L1translate}})}$ (l1M52 maddr l2e) \$\sim$\ (l1off maddr l2e)  $\mapsto_{a}$ {q4} l1e
    \end{lstlisting}
  \item Page Address Level Translation: Final computed physical page-address ($\textsf{addr\_L1}(\vaddr$,$\textsf{l1e}$)) points-to the value stored in the address ($\vpage$).
   \begin{lstlisting}[language=Coq]
    $\hbox{(\TirNameStyle{{PageLevelAccess}})} \qquad$ addr_L1($\vaddr,\textsf{l1e}$) $\mapsto_{p} \vpage$.
    \end{lstlisting} 
\end{itemize}
